Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-04-05 pubs/tfp16/paper.tex
sync just paper

\keywords{strict pure functional programming, automatic parallelization, Haskell, STG,
transparent fault tolerance.}

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade~\cite{Exascale-roadmap} a highly
respected member of the HPC community recently stated publicly that ``Fortran
is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
  Japanese, and international groups are also anticipating the realization
  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are reportedly in the same state.}  C++ has become more common
for new code starts, but their nature remains the same: huge codes with fair
to non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be poor and is explained by
Amdahl's law because the parallel model \emph{as typically implemented} is
bulk synchronous.

\subsection{The trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjunction
with the International Conference on Functional Programming, being now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of these ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that the entire block be brought into scope.  It is almost
ironic, then, that of the mainstream high-performance scientific programming
languages (Fortran, C, and C++), Fortran~95 was the first to introduce the
\texttt{pure} function qualifier, and with enforcement by the compiler.
Otherwise it is entirely up to the programmer to make functions composable and
thread safe.  Here at LANL the use of pure functions (among other techniques
for reducing global data dependencies) is being used to good effect to clean
up some legacy code bases.

@@

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practitioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
of \emph{tying the knot}~\cite{tying-the-knot}.

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get communication (e.g.\ MPI) and IO to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of execution, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece (whether
they've read it or not)~\cite{Hughes89}, they would also be happy to write in
a slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

@@

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model typically implemented with the Message Passing Interface
(MPI)~\cite{MPI}.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  A function in a pure functional language
could be regarded as the essence of the concept of task.  All of these
systems are of course in languages with strict function-call semantics.

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various points in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can be (manually) restarted from the
latest stored state.  Most implementations are explicit, though there has been
some success with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the
programmer.  Ericsson's quasi-functional language Erlang and its runtime
system were designed for fault tolerance~\cite{Cesarini:2009}.  It has long
been recognized that the pure parts of a functional program (excluding, e.g.,
the I/O monad in Haskell) could be safely restarted.  The recent work of
Stewart impressively demonstrates this on a large-scale distributed-memory
system~\cite{Stewart:2013}.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating GHC language, type
system, and runtime features, or significant code optimization, would simply
divert us from our central goals.\footnote{We estimate that we will have, to
  within an order of magnitude, one person-hour of total available development
  time for each person-year that GHC currently represents, or put another way,
  one second for each hour, respectively.}  For another, it means that we are
not undertaking significant morphing of GHC itself: because this is in large
part a student project it must be kept as simple as reasonably possible in
various respects.  Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}

For feasibility, we restrict ourselves to shared-memory implicit parallelism.
The requirement for tail calling is not primarily for efficiency, rather it
is for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, makes this an interesting
exercise that we discuss at some length.

@@

\section{Design and Implementation}
Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are 1) parsing and type checking; 2) transformation to the GHC
\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; 3)
analysis and transformation on Core; 4) transformation of Core to the
\emph{STG} intermediate representation~\cite{PJ-stockhardware}; 5)
transformation of STG to machine code via one of various imperative
representations (Cmm, LLVM, C).

In short, our plan is to use GHC as our front-end to generate one of its
intermediate forms, STG or Core, then escape to our system from there.  There
are various possible ways to accomplish the this: using GHC's option to dump
STG or Core as text, hacking GHC itself and, optionally, using GHC as a
library~\cite{ghc-as-library}.  For now we start with our own STG, with
bridging the gap with GHC scheduled to commence soon.

@@

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters, thus \texttt{id x = x} has by definition arity 1
regardless of the type of the argument (e.g., some function type) it might be
applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
a syntactic notion, from actually calling a function. Thus \texttt{id id x} is
an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
(user- or system- defined function) when actually called is given exactly one
argument (the first one).

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments previously provided.  If the arity is greater than two a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two the underlying
\emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
\emph{PAP}) and new arguments \texttt{x} and \texttt{y}.  If the arity is less
than two (it must be at least one, so one in this case) the underlying
\emph{FUN} is called-with-return with any previous arguments (in case of
\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
remaining argument \texttt{y}.

Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
both of those arguments forced in subsequent evaluation (assuming that they
weren't forced in the evaluation of \texttt{f} itself).

@@

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict, fully strict---all arguments
in an application are evaluated---and the intermediate in which only those
arguments needed by a function call are evaluated.  The STG machinery makes
the choice of evaluation strategy easy to change, both in terms of function
application and building a constructor in the heap.

The other dimension we vary is whether or not constructors are strict.  Other
dimensions are possible, for example whether \texttt{let} bindings are strict,
but we restrict ourselves to function/application and constructor strictness.

@@

All continuations contain a code pointer, optionally some data, and layout
information.  These continuations are maintained on a stack.  When the
evaluation of an object is completed its return is by jumping through the code
pointer of the continuation on the top of the stack (``returning through the
continuation stack'').  For example, the argument frame for a function is one
type of continuation and its associated code simply pops the continuation off
the stack and returns through the next continuation on the stack.

@@

Now consider the evaluation of \texttt{case e of x \{ alts \}} if call-with-return
is disallowed.  If the compiler cannot determine that evaluating \texttt{e}
does not jump somewhere then there must be some mechanism to get to the
\texttt{alts}.  Our solution follows \emph{EA}, using a separate C function
for the \texttt{alts}, and before executing the code to evaluate \texttt{e},
to push a continuation---the address of this function and the free variables
of the \texttt{alts}---onto the stack.  Whatever the code for \texttt{e} might
do, ultimately (in the absence of error or non-termination) the value of
\texttt{e} is in \emph{CV}, and some code somewhere reaches its end and returns
through the continuation stack, thus entering the code for \texttt{alts},
which immediately binds \emph{x} to the value in \emph{CV}.

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  This family is parameterized by the number of actual
arguments in the application and whether each is boxed or unboxed.  Because
we have an explicit stack we are able to use a single C \emph{stgApply} function
taking a contiguous block of machine words (unboxed values or
pointers) on the explicit stack.  Some modifications are required, however,
because \emph{EA} uses call-with-return.  Specifically, when a \emph{FUN} or is
\emph{PAP} applied to more than its arity number of arguments
a call-with-return to the (underlying) \emph{FUN} is made with the arity
number of arguments, then the appropriate \emph{stgApply} function is jumped
to, passing the newly returned object and the remaining arguments.  Our
solution to this entails passing \emph{stgApply} the object to be applied in
\emph{CV}, together with an extra argument indicating which of the object
to be applied and its boxed arguments have been already been evaluated (by
\emph{stgApply}).  If the object to be applied, or some argument, remains to
be evaluated (because of the strictness level), it is jumped to, with
return as ever through the continuation stack.  This is actually fairly
efficient because we can arrange that the stack frame for \emph{stgApply}
is reused, so except for the indirect jumps on return this is essentially
a loop.  For the cases of arguments in excess of the arity of the object
being applied, in effect \emph{stgApply}'s argument frame is popped and
two continuations are pushed, first one to jump to \emph{stgApply} with
the excess arguments, and the next an argument frame for the object to
applied, then the code for the object to be applied is jumped to.

@@

Mutual exclusion for garbage collector--heap check with every heap allocation.
In fact garbage collector can be invoked at any point in time except when
populating a heap object immediately after its allocation.

@@

Various papers describing some of the inner workings of GHC at various point
of its evolution have greatly informed our approach and these have all been
previously cited.

\subsubsection{\emph{Strict but still pure} is not new.}
Very simple implementations have been prototyped, for example
Sheard's.~\cite{Sheard:2003}
%
A plug-in for GHC (prior to 8.0) exists for making functions, but not data
structures, strict.~\cite{strict-ghc-plugin}.  The transformation is on GHC
Core, recursively expand non-recursive let bindings into case bindings.  Thus
it seems that while it makes functions denotationally strict, function arguments
are still evaluated after function call.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order functional language (or family of languages) compiler
that has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.
%
NVIDIA's NOVA language, and corresponding compiler, implements a strict
(``call by value''), pure, polymorphic, higher-order functional language with
built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
\emph{scan}).~\cite{NVIDIA:2013} Three back-ends were implemented: for
sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
targeting NVIDIA GPUs.

\subsubsection{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware.~\cite{Liu:2013,Petersen:2013} As in our
(planned) approach they use GHC as the front end, but exit the GHC toolchain
at the final GHC Core stage, just before transformation to STG.  The essence
of their effort is to target an existing, highly optimizing, Intel functional
language compiler that is ``largely language agnostic'' and performs (among
many other optimization) SIMD vectorization.

In his BSc thesis, Don Stewart described an implementation wherein GHC was
used as a Haskell compiler front-end, dumping STG code as text, parsing that
text, then translating to code for execution on the Java Virtual
Machine.~\cite{Stewart-BSc}.

\subsubsection{Fault tolerance for functional programming.}
Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
distributed Haskell based on distributed exception
handling.~\cite{Pointon:2001}.  However, it appears that their scheme for
handling failed processing elements correctly~\cite{} may have never been
implemented.

@@

Currently we are developing the parallel runtime.  This summer (2016) a
returning student will (partially) bridge the gap between GHC STG and our STG,
and augment our runtime as needed in this effort.  Just as in the Intel Haskell
Research Compiler effort, we will not

>>>>

\keywords{Strict pure functional programming, automatic parallelization,
  Haskell, STG, GHC, GHC Core, transparent fault tolerance.}

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade~\cite{Exascale-roadmap} a highly
respected member of the HPC community recently stated publicly that ``Fortran
is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
  Japanese, and international groups are also anticipating the realization
  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are reportedly in the same state.}  C++ has become more common
for new code starts, but their nature remains the same: huge codes with fair
to non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be good when the parallel model
is bulk synchronous; tasking models are generally avoided unless abstracted
by, e.g., a C++ template library.

\subsection{A trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjunction
with the International Conference on Functional Programming, being now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that entire blocks be brought into scope.  It is almost ironic,
then, that of the mainstream high-performance scientific programming languages
(Fortran, C, and C++), Fortran~95 was the first to introduce the \texttt{pure}
function qualifier, and with enforcement by the compiler.  Otherwise it is
entirely up to the programmer to make functions composable and thread safe.
Here at LANL the use of pure functions (among other techniques for reducing
global data dependencies) is being used to good effect to clean up some legacy
code bases.

@@

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practitioners who are developing an appreciation for pure functional
programming in Haskell.  Still, they see their needs as different than other
Haskell users: performance is paramount, vectors (``Efficient
Arrays''~\cite{vector}) are a (perhaps \emph{the}) primary data structure, and
non-strictness or laziness is more of a nuisance---even a serious
hindrance---than a convenience, much less ever algorithmically essential in
the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's general attribute
grammars~\cite{Johnsson87}, or other instances of \emph{tying the
  knot}~\cite{tying-the-knot}.

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get I/O and inter-process communication to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of execution, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece (whether
they've read it or not)~\cite{Hughes89}, they would also be happy to write in
a slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

@@

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model typically implemented with the Message Passing Interface
(MPI) \cite{MPI}.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  A function in a pure functional language
could be regarded as the essence of the concept of task.  All of these
systems are of course in languages with strict function-call semantics.

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various points in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can be (manually) restarted from the
last stored state.  Most implementations are explicit, though there has been
some success with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the
programmer.  Ericsson's quasi-functional language Erlang and its runtime
system were designed for fault tolerance~\cite{Cesarini:2009}.  It has long
been recognized that the pure parts of a functional program (excluding, e.g.,
the I/O monad in Haskell) could be safely restarted.  The recent work of
Stewart impressively demonstrates this, in a pure functional setting, on a
large-scale distributed-memory system~\cite{Stewart:2013}.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating GHC language, type
system, and runtime features, or significant code optimization, would simply
divert us from our central goals.\footnote{We estimate that we will have, to
  within an order of magnitude, one person-hour of total available development
  time for each person-year that GHC currently represents, or put another way,
  one second for each hour, respectively.}  For another, it means that we are
not undertaking significant morphing of GHC itself: because this is in large
part a student project it must be kept as simple as reasonably possible in
various respects.  Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}

For feasibility, we restrict ourselves to shared-memory implicit parallelism.
The requirement for tail calling is not primarily for efficiency, rather it
is for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, makes this an interesting
exercise that we will discuss at some length.

@@

\section{Design and Implementation}
Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are 1) parsing and type checking; 2) transformation to the GHC
\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; 3)
analysis and transformation on Core; 4) transformation of Core to the
\emph{STG} intermediate representation~\cite{PJ-stockhardware}; 5)
transformation of STG to machine code via various imperative
representations (Cmm, LLVM, C, native code generation).

In short, our plan is to use GHC as our front-end to generate one of its
intermediate forms, STG or Core, then escape to our system from there.  There
are various possible ways to accomplish the this: using GHC's option to dump
STG or Core as text, hacking GHC itself and, optionally, using GHC as a
library~\cite{ghc-as-library}.  For now we start with our own STG, with
bridging the gap with GHC scheduled to commence quite soon.

@@

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters in a function definition, thus \texttt{id x = x}
has by definition arity one regardless of the type of the argument (e.g., some
function type) it might be applied to.  We distinguish \emph{application},
e.g., \texttt{f x y}, which is a syntactic notion, from actually calling a
function. Thus \texttt{id id x} is an application of \texttt{id} to two
arguments, but the underlying \emph{FUN} (user- or system- defined function)
when actually called is given exactly one argument (the first one).

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} or \emph{PAP} (partial application of a \emph{FUN}) is
determined---in the case of \emph{PAP} the arity is that of the underlying
\emph{FUN} less the number of arguments previously provided.  If the arity is
greater than two a new \emph{PAP} is created with arguments \texttt{x} and
\texttt{y} inserted and the new \emph{PAP} returned.  If the arity is equal to
two the underlying \emph{FUN} is tail-called (jumped to) with any previous
arguments (in case of \emph{PAP}) and new arguments \texttt{x} and \texttt{y}.
If the arity is less than two (it must be at least one, so one in this case)
the underlying \emph{FUN} is called with any previous arguments (in case of
\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
remaining argument \texttt{y}.

Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
both of those arguments possibly forced in subsequent evaluation (assuming
that they weren't forced in the evaluation of \texttt{f} itself).

@@

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict; function strict, in which only
those arguments needed by a function call are evaluated; and application
strict, in which all arguments of an application are evaluated.  The other
dimension we vary is whether or not constructors are strict.  Other dimensions
are possible, for example whether \texttt{let} bindings are strict, but we
restrict ourselves to function/application and constructor strictness.  The
STG machinery makes the choice of evaluation strategy easy to change, both in
terms of function application and building a constructor in the heap.

@@

Continuations contain a code pointer, optionally some data, and layout
information.  These continuations are maintained on a stack.  When the
evaluation of an object is completed its return is by jumping through the code
pointer of the continuation on the top of the stack (``returning through the
continuation stack'').  For example, the argument frame for a function is one
type of continuation and its associated code simply pops the continuation off
the stack and returns through the next continuation on the stack.

@@

Now consider the evaluation of \texttt{case e of x \{ alts \}} if
call-with-return is disallowed.  If the compiler cannot determine that
evaluating \texttt{e} does not jump somewhere then there must be some
mechanism to get to the \texttt{alts}.  Our solution follows \emph{EA}, using
a separate C function for the \texttt{alts}, and before executing the code to
evaluate \texttt{e}, to push a continuation---the address of this function and
the free variables of the \texttt{alts}---onto the stack.  Whatever the code
for \texttt{e} might do, ultimately (in the absence of error or
non-termination) the value of \texttt{e} is in \emph{CV}, and some code
somewhere has reached its end and returns through the continuation stack, thus
entering the code for \texttt{alts}, which immediately associates \texttt{x}
with the value in \emph{CV}.

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  This family is parameterized by the number of actual
arguments in the application and whether each is boxed or unboxed.  Because we
have an explicit stack we are able to use a single C \emph{stgApply} function
taking a contiguous block of machine words (unboxed values or pointers) on the
explicit stack.  Some modifications are required, however, because \emph{EA}
uses call-with-return.  Specifically, when a \emph{FUN} or is \emph{PAP}
applied to more than its arity number of arguments a call-with-return to the
(underlying) \emph{FUN} is made with the arity number of arguments, then the
appropriate \emph{stgApply} function is jumped to, passing the newly returned
object and the remaining arguments.  Call-with-return becomes even more
convenient for strict evaluation:  arguments can be evaluated in a tight
loop.

Our solution to this entails passing \emph{stgApply} the object to be applied
in \emph{CV}, together with an extra argument---call it
\emph{index}---indicating which of the object to be applied and its boxed
arguments have been already been evaluated (previously by \emph{stgApply}).
If the object to be applied, or some argument (because of the strictness
level), remains to be evaluated, it is jumped to, with return as ever through
the continuation stack.  This is actually fairly efficient because we can
arrange that the stack frame for \emph{stgApply} is reused, updating only
\emph{index}, so except for the indirect jumps on return this is essentially a
loop.  For the cases of arguments in excess of the arity of the object being
applied, in effect \emph{stgApply}'s argument frame is popped and two
continuations are pushed, the first one to jump to \emph{stgApply} with the
excess arguments, the next an argument frame for the object to applied, then
the code for the object to be applied is jumped to.

@@

%Mutual exclusion for garbage collector--heap check with every heap allocation.
%In fact garbage collector can be invoked at any point in time except when
%populating a heap object immediately after its allocation.

@@

Various papers describing some of the inner workings of GHC at various points
in its evolution have greatly informed our approach and these have all been
previously cited.

\subsubsection{\emph{Strict but still pure} is not new.}
Very simple implementations have been prototyped, for example
Sheard's.~\cite{Sheard:2003}
%
A plug-in for GHC (prior to 8.0) exists for making functions, but not data
structures, strict using a transformation on GHC Core that recursively expands
non-recursive let bindings into case bindings.~\cite{strict-ghc-plugin}.  Thus
it seems that while it makes functions denotationally strict, function
arguments are still evaluated after function call.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order functional language (or family of languages) compiler
that has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.
%
NVIDIA's NOVA language, and corresponding compiler, implements a strict
(``call by value''), pure, polymorphic, higher-order functional language with
built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
targeting NVIDIA GPUs.

\subsubsection{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
planned approach they use GHC as the front end, but exit the GHC toolchain
at the final GHC Core stage, just before transformation to STG.  The essence
of their effort is to target an existing, highly optimizing, Intel functional
language compiler that is ``largely language agnostic'' and performs (among
many other optimization) SIMD vectorization.

Don Stewart described an implementation wherein GHC was used as a Haskell compiler
front-end, dumping STG code as text, parsing that text, then translating to
code for execution on the Java Virtual Machine.~\cite{Stewart-BSc}.

\subsubsection{Fault tolerance for functional programming.}
Pointon et al.\ developed a limited fault tolerance mechanism for Glasgow
distributed Haskell based on distributed exception
handling~\cite{Pointon:2001}.  However, it appears that their scheme for
handling failed processing elements correctly~\cite{} may have never been
implemented.

@@

Currently we are developing the parallel runtime.  This summer (2016) a
returning student will (partially) bridge the gap between GHC STG (or Core)
and our STG, and augment our runtime as needed in this effort.  Just as in the
Intel Haskell Research Compiler effort, we will not attempt to reuse the GHC
runtime.
